---
title: "Package crsim: making consumer-resource simulations easier"
author: "Andrew D. Letten"
output: 
#  rmarkdown::html_document: default
  rmarkdown::pdf_document: 
#    number_sections: true
vignette: >
  %\VignetteIndexEntry{crsim, how to...}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`crsim` is an R package for defining ode models of consumer-resource interactions for use alongside `deSolve`. It also provides a number of additional functions for visualising functional responses and output dynamics from `deSolve`.

The main user function in `crsim` is `make_par_list()`, which facilitates the definition and parameterisation of a desired model. Features/options include: 

- Number of consumers/resources
- Consumer functional response (type I or type II)
- Resource dynamic (chemostat, logistic and/or pulsed)
- Resource type (substitutable or essential)
- Time dependent consumption parameters 

See `?make_par_list` for all argument options.

**Examples demonstrated in this vignette include:**

- One type 2 consumer and one logistically growing resource
- One type 2 consumer and one continuously supplied resource (i.e. chemostat) 
- Two type 2 consumers and one logistically growing resource
- Two type 1 consumers and two substitutable resources in a chemostat
- Two type 2 consumers and one externally pulsed resource 
    * Continuous mortality
    * Serial batch culture
- Two type 2 consumers with time dependent consumption parameters and one continuously supplied resource


### Type 1 consumer on a logistically growing resource


```{r setup}
library(crsim)
library(deSolve)
```

```{r}
pars <- make_par_list(
  spnum = 1, 
  resnum = 1,
  linear = FALSE,
  mumatrix = list(matrix(0.1)),
  resspeed = 3,
  resconc = 0.2
)
```

Visualise funcitonal responses with `crsim::plot_funcresp`. Faceted by resources and time dependent parameterisations (when relevant).

```{r, fig.width = 3, fig.height = 3}
plot_funcresp(pars, maxx = 0.2)
```

The function `time_vals` is used for setting total simulation time (and is used to set the resource pulse interval when used in conjunction with an event function [see below]). The function `initiate_state` sets the starting values of state variables defaulting to 10 for consumers and the resource supply concentration for resources.

```{r}
happenings <- time_vals(total = 500)

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda"
  )
```

`plot_cr_sim` plots the output dynamics.

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```

### Type 1 consumer on a continuously suppplied resource

To switch to a continuously supplied resource,use `chemo = TRUE'.

```{r}
pars <- make_par_list(
  spnum = 1, 
  resnum = 1,
  linear = FALSE,
  mumatrix = list(matrix(0.1)),
  resspeed = 3,
  resconc = 0.2,
  chemo = TRUE
)
```

```{r}
happenings <- time_vals(total = 500)

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda"
  )
```

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```

### Two type 2 consumers on a logistically growing resource

```{r}
pars <- make_par_list(
  spnum = 2, 
  resnum = 1,
  linear = FALSE,
  mumatrix = list(matrix(c(0.7,0.05), 
                    nrow = 2, 
                    ncol = 1,
                    byrow = TRUE)),
  kmatrix = matrix(c(2, 0.015), 
                   nrow = 2, 
                   ncol = 1, 
                   byrow = TRUE),  
  resspeed = 3,
  resconc = 0.2
)
```

```{r, fig.width = 3, fig.height = 3}
plot_funcresp(pars, maxx = 0.2)
```

```{r}
happenings <- time_vals(total = 2000)

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda"
  )
```

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```
Owing to the trade-off in functional responses (gleaner-opportunist) and resource fluctuations generated by N2, the two consumers are able to coexist via the coexistence mechanism known as 'relative nonlinearity of competition'.

### Two type 1 consumers and two substitutable resources in a chemostat

```{r}
pars <- make_par_list(
  spnum = 2, 
  resnum = 2,
  linear = TRUE,
  mumatrix = list(matrix(c(0.07,0.03,
                           0.04,0.05), 
                    nrow = 2, 
                    ncol = 2,
                    byrow = TRUE)),
  resspeed = 3,
  resconc = 0.2,
  chemo = TRUE,
  essential = FALSE
)
```

```{r, fig.width = 6, fig.height = 3}
plot_funcresp(pars, maxx = 0.2)
```

```{r}
happenings <- time_vals(total = 1000)

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda"
  )
```

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```
Resource partitioning, where each consumer is a better competitor (lower R*) for a different resource, results in coexistence.

### Two type 2 consumers and one externally pulsed resource (continuous mortality)

External resource pulsing is handled using an event function, `crsim::eventfun_respulse` which is passed in a list to `deSolve::ode`. The size of the resource pulse is specified as a numeric with the argument `respulse` in the call to `make_par_list`. To prevent any other mode of resource supply, `resspeed` should be set to zero. The frequency of pulsing is set with the `pulse` argument in `crsim::time_vals`. The resultant vector of pusling intervals should be included in the same list containing the event function.

```{r}
pars <- make_par_list(
  spnum = 2, 
  resnum = 1,
  linear = FALSE,
  mumatrix = list(matrix(c(0.7,0.05), 
                    nrow = 2, 
                    ncol = 1,
                    byrow = TRUE)),
  kmatrix = matrix(c(2, 0.015), 
                   nrow = 2, 
                   ncol = 1, 
                   byrow = TRUE),  
  resspeed = 0, # set to zero for no additional resource supply 
  resconc = 0.2,
  respulse = 0.3 # resource pulse size
)
```

```{r, fig.width = 3, fig.height = 3}
plot_funcresp(pars, maxx = max(pars$respulse))
```

```{r}
happenings <- time_vals(total = 2000, pulse = 100) # 'pulse' sets the interval 

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda",
  # events argument takes a list including the event function and a vector of pulsing intervals
  events = list(func = eventfun_respulse, time = happenings$pulseseq),
  )
```

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```

### Two type 2 consumers and one externally pulsed resource (serial transfer with pulsed mortality)

*To be implemented*

### Two type 2 consumers with time dependent consumption parameters and one continuously supplied resource

Using time dependent consumption parameters is appropriate for capturing basic consumer-environment interactions (e.g. differential responses to temperature, pH, antibiotics etc.). Time varying parameters can be implemented in `deSolve` with forcing functions. `crsim` provides a straightforward means of implementing forcing functions in the background. Three additional arguments are needed in `make_par_list`: `timepars`, `timeswitch` and `time_switch_length`. `timepars` should be set to `TRUE`; `timeswitch` adjusts the frequency of switching between 'environmental states'; `time_switch_length` adjust the total time over which the environment varies (for most applications this should equal the total time in the call to `time_vals`). 

In addition, if `time_pars` = TRUE, the mumatrix argument `make_par_list` expects a list containing two matrices, reflecting two different environmental states. At this stage `crsim` only handles two states with equal time allotted to each state.

```{r}
pars <- make_par_list(
  spnum = 2, 
  resnum = 1,
  linear = FALSE,
  mumatrix = list(matrix(c(0.3,0.1), 
                    nrow = 2, 
                    ncol = 1,
                    byrow = TRUE),
                  matrix(c(0.2,0.5), 
                    nrow = 2, 
                    ncol = 1,
                    byrow = TRUE)),
  chemo = TRUE,
  resspeed = 0.1,
  resconc = 0.1,
  timepars = TRUE,
  timeswitch = 50,
  timeswitch_length = 2000
)
```

```{r, fig.width = 3, fig.height = 6}
plot_funcresp(pars, maxx = max(pars$resconc))
```

```{r}
happenings <- time_vals(total = 2000)

m1 <- ode(
  func = def_cr_ode,
  y = initiate_state(vars=pars),
  parms = pars,
  times = happenings$totaltime,
  method = "lsoda",
  )
```

```{r, fig.width = 6, fig.height = 3}
plot_crsim(m1, pars) 
```

The two consumers coexist due species-specific environmental responses leading to a temporal storage effect (formalization of temporal niche paritioning arising from density dependence in environment-competition covariance).
